---
title: 'Introduction to Computational Statistics with R'
subtitle: "Stats 102A"
author: "Miles Chen, PhD"
date: 'Week 1 Monday'
footer: Copyright Miles Chen, 2019. Please do not distribute or post without permission.
output: ioslides_presentation
widescreen: TRUE
---

## Welcome!

- My name is Miles Chen
- mileschen@stat.ucla.edu
- http://ccle.ucla.edu
- Office: Math Sciences 8105
- Office Hours: Mondays, Tuesdays, Thursdays 2-2:50 PM, and by appointment

## Personal note about office hours:

I **like** when students come to office hours with questions about material. I love to explain things and to help students understand.

I **like** when students come to office hours to tell me more about themselves and to seek counsel about classes to take or next steps. I am also happy to help students who face difficult circumstances and may need some exceptions to be made. Please do not hesistate to visit office hours.

I am happy to correct mistakes that may have been made. I don't enjoy getting in arguments with students over homework points. I don't like arguing whether a particular mistake should be a 5 or 10 point deduction.

## My commitment to you

- You are the paying customer. It is my goal that you are completely satisfied with the course and with the teaching I provide to you. I hope to receive very high reviews from you at the end of the quarter.

- My job is to educate you to learn the skills and methods for data analysis in the context of computational statistics.

- I love the subject of statistics, and I love teaching. I hope to demonstrate this to you in my lectures. For those of you who are stats majors/minors, I hope you will be convinced that you have made a good choice.

## My expectations of you

- If you are not happy with the class, please let me know so I can fix the problem. Please do not grumble quietly and leave me a bad review.

- I expect you to put forth a good faith effort to learn the topics covered in the course.

- I expect you to wrestle with difficult topics and not give up at the first sign of resistance.


# Let's Get Started

# What is this class about?

## Big Picture: Statistics

## Intro Stats

### Why do we study statistics? 

- To gain a better understanding of the world. (this could be said about almost any subject)
- The subject of statistics allows us to take data and then make meaningful conclusions from it.

### "What does this data tell me about the world?"

- With data, we can describe and summarize it.
- With data, we can ask, "could the results I have be a random fluke, or does it indicate something special?"

- In your intro stats class, you learned descriptive statistics, and then moved on to inferential statistics where you made confidence intervals and performed hypothesis tests.

- Confidence intervals and hypothesis tests depend on the Central Limit Theorem.

## Big Picture: Hypothesis Testing {.build}
- When you made a conclusion at the end of a hypothesis test, you were essentially saying something along the lines of:

- I hypothesize that the population has a mean of 10. The laws of probability tell me that if the population has a mean of 10, then samples chosen randomly from the population should have a mean close to 10.

- The sample I have has a mean of 11.3. According to the laws of probability (CLT) and the variance I observed in my sample, there is a very low probability (p-value) that a random sample drawn from my hypothesized population will have a mean of 11.3. 

- Therefore, I do not believe my randomly selected sample comes from a population with a mean of 10.

## Models and Analytic Approaches

- The hypothesis testing performed in intro stats is an example of a simple data-generating model.

- In our "model of reality," we test the belief that the data we have is generated by a population with a mean of 10 (the model parameter).

- Our model states that samples means should come from a normal distribution.
- Our understanding of probability allows us to test our model in an analytic way.

## This Course: Computational Statistics

- For many models and situations, the math can become extremely complicated or even impossible to solve.
- We may be able to use computational intensive methods (like simulations and resampling) to make inferences about the parameters.
- Computational Statistics is a broad field in which we employ the use of a computer and computationally intensive methods to answer our statistical questions.
- Computational Statistics is not simply "doing statistics with a computer"


## What we'll cover in the course

- Stats 102A is your introduction to the field of computational statistics.

- The course can be thematically split into two parts.
    + Learning the tools of computational statistics, and how they work (R and its packages)
    + Introduction to the methods of computational statistics

- I liken this arrangement to a woodshop class. In the beginning you learn the tools. Learning tools is not actually woodworking. However, once you learn the tools, then you can begin to work on your craft.

## What we'll cover: Tools

1) Week 1: R Data structures, subsetting
2) Week 2: Flow control, writing functions
3) Week 3: Scope, Object-oriented programming, 
4) Week 4: Packages readr, tidyr, dplyr, web scraping
5) Week 5: Regular expressions, ggplot2, Midterm Exam

## What we'll cover: Computational Statistics

6) Week 6: Numerical methods: floating point arithmetic, root finding
7) Week 7: Numerical optimization
8) Week 8: Random distributions in R and their use in simulations
9) Week 9: Randomization tests, permutation tests, bootstrapping
10) Week 10: More bootstrapping, kernel density estimation

## Data Science

Data Science Venn Diagram:

- Coding Skills
- Math & Stats Knowledge
- Domain Expertise

## Coding

- Expect to do a lot of coding in this class

## R Markdown

- R Markdown integrates the R code and analysis into your document.

- Changes made to the data or anywhere in the analysis process are updated directly in the document.

- Documents can be "rendered" into articles or slides in HTML or PDF

- This presentation is made with R Markdown. 

## A quick note about Markdown 

- Markdown was created with readability in mind. This means that the plain text file should still be readable even if it hasn't been rendered.

- Lists look like ordered lists. Emphasized text looks emphasized.

- Block quotes, code chunks, and tables resemble the final output even in raw text.

- The downside is that you don't have as much fine control over the *exact* styling or placement.

## Resources

- http://rmarkdown.rstudio.com/

- Cheat sheet: https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf

## Getting Started

RStudio should auto-install the necessary packages when you create a new R Markdown document. But you can also install the package yourself.

```{r, eval=FALSE}
install.packages("rmarkdown")
```

The new document will contain a YAML header, which you should modify. Write your document below the header.

```c
---
title: "Untitled"
author: "Put Your Name Here"
date: "January 7, 2019"
output: html_document
---
```

## Rendering the document

You can change the output format in the YAML header. Options include: `html_document`, `pdf_document`, `word_document`, as well as presentation formats like `ioslides_presentation`, which is currently being used for this presentation.

When you are ready to render the document, click the **Knit** button, or enter `Ctrl + Shift + K`

## Troubleshooting R Markdown

Sometimes R Markdown won't render the document.

Usually this is caused by an error in one of the code chunks. R Markdown produces error messages, but in the beginning, it can be difficult to decipher.

One strategy to figure out what is wrong:

1) Create a new blank R Markdown document.
2) Render the document
3) Copy a chunk of text and code from the broken document to the new document.
4) Render it.
5) Repeat steps 3 and 4 until the new document no longer renders properly. You will know that there is an error in the last chunk you copied over.


# Data Structures in R

## Data Structures and Dimensionality

dimension|Homogeneous|Heterogeneous
:---:|:---:|:---:
1d|Atomic Vector|List (generic vector)
2d|Matrix|Data frame
nd|Array

[credit to Hadley Wickham, from whom much of this presentation draws content]

## Atomic Vectors

R has six basic atomic vector types:

typeof() | mode() | storage.mode()
-------|------|----------
logical | logical | logical
double | numeric | double
integer | numeric | integer
character | character | character
complex | complex | complex
raw | raw | raw

We'll only worry about the first four.

## Atomic Vectors

Atomic vectors are homogenous. All objects in the vector must be of the same type.

## Vector Types - Logical

Boolean values `TRUE` or `FALSE`

Always type out the whole word `TRUE` or `FALSE` in all caps. These are reserved by R to always be their logical values. R understands `T` and `F`, but values can be assigned to an object called `T`, which could cause issues if you expect `T` to mean TRUE.

```{r}
typeof(TRUE)
a <- c(FALSE, FALSE, FALSE)
typeof(a)
```

## Vector Types - Character

Character strings

```{r}
typeof("Hello")
b <- c('Adam', 'Bob', 'Chuck')
typeof(b)
```

## Vector Types - Double

Floating point numerical values (default numerical type)

```{r}
typeof(1.33)
typeof(4)
typeof(c(1,2,3))
```

## Vector Types - Integer

Integer values. Indicated with `L` after the number. Also created with the colon :

```{r}
typeof(1L)
typeof(1:3)
```

